---
title: "National Malaria Elimination Program Dashboard: Technical Documentation and User Guide"
author: "Enku Demssie"
date: "June 2026"
editor: visual
page-layout: article

format:
  html:
    toc: true
    toc-depth: 4
    self-contained: true
    number-sections: false
    css: styles.css

execute:
  eval: false
---

# Introduction

The National Malaria Elimination Program (NMEP) malaria dashboard was developed to support routine monitoring and analysis of malaria indicators using data from the national Health Management Information System (HMIS), implemented through DHIS2.

The dashboard integrates routine HMIS data on malaria testing, confirmed cases, species distribution, and malaria-related deaths across inpatient and outpatient services. These data are used for monitoring malaria trends over time and across geographic areas.

The system is designed to enable standardized analysis of key malaria indicators at woreda level, supporting routine program monitoring, reporting, and decision-making.

This document describes the step-by-step process used to develop the dashboard, including data extraction, processing, indicator construction, and deployment.

# Overview of The Dashboard Development Process

The development of the HMIS-based NMEP dashboard followed a structured workflow consisting of the following key steps:

1.  Data extraction from DHIS2
2.  Data preparation and cleaning
    - Data preparation
    - General cleaning and harmonization
    - Geographic cleaning
3.  Dashboard dataset preparation
    - Historical data integration
    - Lookup tables
    - Time enrichment
    - Aggregation
4.  Visualization and dashboard development
5.  Deployment and routine updating

Each step is described in detail in the following sections.

# 1. Data Extraction from DHIS2

Data for the dashboard were extracted from the national Health Management Information System (HMIS) implemented through DHIS2. The extraction covered the period from January 2020 to April 2026 and was conducted at the woreda level, which was selected as the lowest administrative unit for analysis and visualization.

Data were drawn from three primary HMIS datasets to support comprehensive analysis of malaria burden, service delivery, and reporting performance.

## 1.1 Disease surveilance Data

The disease surveillance dataset was extracted from DHIS2 using the following components:

- Data elements (malaria-related indicators)\
- Period\
- Organizational unit\
- Outcome (morbidity and mortality disaggregation)\
- Department (inpatient and outpatient disaggregation)

These components enabled structured disaggregation of malaria burden across time, geography, and service delivery settings.

A total of 16 malaria-related data elements were included. These comprised both ICD-10–based data elements (used prior to mid-2023) and ICD-11–based data elements (currently in use). This approach ensured continuity of analysis despite changes in disease classification systems.

**Disease Surveillance Data Elements**

**ICD-10 (2018–2023)**

- B50–196 Malaria (Plasmodium falciparum malaria)\
- B50–197 Malaria (Plasmodium falciparum with cerebral complications)\
- B50–198 Malaria (Other severe and complicated Plasmodium falciparum malaria)\
- B50–199 Malaria (Plasmodium falciparum unspecified)\
- B50–200 Malaria (Plasmodium vivax malaria)\
- B50–201 Malaria (Plasmodium vivax with other complications)\
- B50–202 Malaria (Plasmodium vivax without complication)\
- B50–203 Mixed malaria (Other parasitologically confirmed malaria)\
- B50–204 Malaria (Unspecified malaria)

**ICD-11 (2023–present)**

- 1F40 Malaria due to *Plasmodium falciparum*\

- 1F41 Malaria due to *Plasmodium vivax*\

- 1F42 Malaria due to *Plasmodium malariae*\

- 1F43 Malaria due to *Plasmodium ovale*\

- 1F44 Other parasitologically confirmed malaria\

- 1F40/1F41 Mixed malaria (*P. falciparum* and *P. vivax*)\

- 1F45 Malaria without parasitological confirmation

## 1.2 Service Delivery Data

Service delivery data were extracted separately from DHIS2, as this dataset does not support the additional disaggregation dimensions used in the disease surveillance dataset (i.e., outcome and department). As a result, only the standard DHIS2 components were applied:

- Data elements\
- Period\
- Organizational unit

The dataset included three key malaria-related data elements:

- Malaria tests (legacy indicator used prior to 2023: “Total number of slides or RDT performed for malaria diagnosis”)\
- Malaria tests (current indicator: “Slides or RDT performed for malaria diagnosis”)\
- Confirmed positive malaria tests

This separation ensured consistency with the DHIS2 service reporting structure while maintaining data integrity across indicator definitions and system updates.

## 1.3 Reporting Completeness and Timeliness Data

Data on service report completeness and timeliness were extracted from the integrated service reporting dataset used across malaria, NTD, and NCD programs, as no malaria-specific reporting performance dataset is available.

The dataset included nine data elements, disaggregated by facility type:

- Hospitals, health centers, and private facilities\
- Basic health posts\
- Comprehensive health posts

For each facility type, the following indicators were extracted:

- Actual reports submitted\
- Reports submitted on time\
- Expected reports

These data elements were used to assess service report completeness and timeliness across different levels of the health system.

**Data Export Format**

All datasets were manually exported from DHIS2 in CSV format. These exports formed the input datasets for subsequent data cleaning, transformation, and indicator calculation within the dashboard development workflow.

# 2. Data Preparation & cleaning

The raw HMIS datasets (disease surveillance, service delivery, and reporting completeness & timeliness) were processed to create a unified analysis-ready dataset for the dashboard. The cleaning and transformation process was implemented in R.

## 2.1 Data preparation

### 2.1.1 Standardization of variable names and geographic fields

The three datasets extracted from DHIS2 were first read into the R environment.

```{r}
library(tidyverse) 
library(lubridate)

# reading the disease report dat
df_species <- suppressWarnings(
  read_csv("data/raw/April 2026 hmis/disease_dat_april2026.csv", show_col_types = FALSE))

# reading the service report data
df_service <- suppressWarnings(read_csv("data/raw/April 2026 hmis/service_dat_april2026.csv", show_col_types = FALSE))

# reading the dataset for service report completeness & timeliness
df_report <-suppressMessages(
  suppressWarnings(
  read_csv("data/raw/April 2026 hmis/report_dat_april2026.csv",show_col_types = FALSE)))

# checking the variable names for the 3 dataset
names(df_species)
names(df_service)
names(df_report)

# preparing a function to rename the adminstartive levels 
rename_geo <- function(df){
  df |>
    rename(
      period = periodname,
      region = orgunitlevel2,
      zone   = orgunitlevel3,
      woreda = orgunitlevel4
    )
}

# performing demonstrative leve renaming using the above function
df_species <- rename_geo(df_species)
df_service <- rename_geo(df_service)
df_report  <- rename_geo(df_report)

```

At this stage, variable names and administrative unit fields were standardized across all datasets to ensure consistency prior to integration.

### 2.1.2 Reshaping to long format

Each dataset was transformed from wide format to long format to enable indicator-level analysis.

```{r}
df_species_long <- df_species |>
  pivot_longer(cols = matches("ESV|B50"),
               names_to = "data_type",
               values_to = "value")

df_service_long <- df_service |>
  pivot_longer(cols = contains("RDT"),
               names_to = "data_type",
               values_to = "value")

df_report_long <- df_report |>
  pivot_longer(cols = matches("NCD|NTD"),
               names_to = "data_type",
               values_to = "value")
```

### 2.1.3 Standardization of indicator dimensions

To allow integration, missing categorical variables were added to ensure a consistent structure across datasets.

```{r}
df_service_long <- df_service_long |>
  mutate(department = NA_character_,
         outcome = NA_character_,
         facility_type = NA_character_)

df_report_long <- df_report_long |>
  mutate(department = NA_character_,
         outcome = NA_character_)

df_species_long <- df_species_long |>
  mutate(facility_type = NA_character_)
```

### 2.1.4 Dataset integration

After standardizing variable names and aligning dataset structures, the disease surveillance, service delivery, and reporting datasets were integrated into a single unified dataset.

```{r}
combined_hmis <- bind_rows(
  df_species_long,
  df_service_long,
  df_report_long
)
```

The resulting dataset (\`combined_hmis\`) serves as the primary integrated dataset for subsequent processing steps, including geographic cleaning and indicator construction for the dashboard.

```{r}
# saving the integrated dataset
saveRDS(combined_hmis,
        "data/raw/April 2026 hmis/combined-hmis-apr_2026.rds")
```

## 2.2. General Data Cleaning and Harmonization

Following the creation of the integrated dataset (`combined_hmis`), a general cleaning and harmonization process was applied to standardize variable names, data elements, time variables, and geographic identifiers prior to geographic cleaning.

This step ensured that the dataset was consistent, structured, and ready for geospatial standardization and downstream analysis.

### 2.2.1 Loading processed dataset

The integrated HMIS dataset from the previous step was loaded into the R environment for further cleaning and transformation.

```{r}
# libraries 
library(tidyverse) 
library(lubridate) 
library(janitor) 
library(sf) 
library(stringr)

#reading the integrated dataset
combined_hmis <- readRDS( "data/raw/April 2026 hmis/combined-hmis-apr_2026.rds" )
```

### 2.2.2 Standardization of variable names and text fields

Variable names and geographic text fields were cleaned to ensure consistency across administrative units.

```{r}
combined_hmis <- combined_hmis |>
  janitor::clean_names() |>
  mutate(
    region = gsub("Region", "", region),
    region = gsub("region", "", region),
    region = gsub("City Administration", "", region),
    region = gsub("al Health Bureau", "", region),
    region = gsub("Ethiopian", "", region),
    region = gsub("Ethiopia", "", region),
    region = gsub("  ", " ", region),
    region = str_trim(region)
  ) |>
  mutate(
    zone = gsub("Health Office", "", zone),
    zone = gsub("Zone", "", zone),
    zone = gsub("City Administration", "", zone),
    zone = gsub("Zone Health Department", "", zone),
    zone = gsub("Health Department", "", zone),
    zone = gsub("Health department", "", zone),
    zone = gsub("Woreda", "", zone),
    zone = gsub("Zonal", "", zone),
    woreda = gsub("Town Adminstration", "", woreda),
    woreda = gsub("Health Office", "", woreda),
    woreda = str_squish(woreda)
  )
```

### 2.2.3 Harmonization of data elements

Malaria-related indicators and reporting variables were standardized into a simplified classification system to enable consistent analysis across ICD-10 and ICD-11 periods.

```{r}
pf_conf <- c("B50-196 Malaria (Plasmodium falciparum malaria)",
             "B50-197 Malaria (Plasmodium falciparum malaria with cerebral complications)",
             "B50-198 Malaria (Other severe and complicated Plasmodium falciparum malaria)",
             "B50-199 Malaria (Plasmodium falciparum malaria unspecified)",
             "ESV-ICD11 1F40 - Malaria due to Plasmodium falciparum",
             "ESV-ICD11 1F44 - Other parasitologically confirmed malaria",
             "B50-203 Mixed Malaria (Other parasitologically confirmed malaria)"
)

pv_conf <- c("B50-200 Malaria (Plasmodium vivax malaria)",
             "B50-201 Malaria (Plasmodium vivax malaria with other complications)",
             "B50-202 Malaria (Plasmodium vivax malaria without complication)",
             "ESV-ICD11 1F41 - Malaria due to Plasmodium vivax"
)

mixed_conf <- c("ESV-ICD11 1F40/1F41 -  Malaria due to Plasmodium falciparum associated with Malaria due to Plasmodium Vivax (Mixed Malaria)",
                "B50-203 Mixed Malaria (Other parasitologically confirmed malaria)"
)

pm_conf <- c("ESV-ICD11 1F42 - Malaria due to Plasmodium malariae")

po_conf <- c("ESV-ICD11 1F43 - Malaria due to Plasmodium ovale")

clinical_malaria <- c("B50-204 Malaria (Unspecified malaria)",
                      "ESV-ICD11 1F45 - Malaria without parasitological confirmation")

tested <- c("MAL_Slides or RDT performed for malaria diagnosis",
            "-2017_Total number of slides or RDT performed for malaria diagnosis")

positives <- c("MAL_Slides or RDT Positive")
actual_reports <- c("08.1 - Malaria, NTD, NCD | Basic Health post | Monthly - Actual reports",
                    "08.2 - Malaria, NTD, NCD | Hospital, Health center, Clinic | Monthly - Actual reports",
                    "08.3 - Malaria, NTD, NCD | Comprehensive HP | Monthly - Actual reports")

actual_on_time <- c("08.1 - Malaria, NTD, NCD | Basic Health post | Monthly - Actual reports on time",
                    "08.2 - Malaria, NTD, NCD | Hospital, Health center, Clinic | Monthly - Actual reports on time",
                    "08.3 - Malaria, NTD, NCD | Comprehensive HP | Monthly - Actual reports on time")

expected <- c("08.1 - Malaria, NTD, NCD | Basic Health post | Monthly - Expected reports",
              "08.2 - Malaria, NTD, NCD | Hospital, Health center, Clinic | Monthly - Expected reports",
              "08.3 - Malaria, NTD, NCD | Comprehensive HP | Monthly - Expected reports")
```

### 2.2.4 Standardization of data elements

All indicators were standardized into simplified categories for analysis.

```{r}
combined_hmis <- combined_hmis |>
  mutate(
    data_type = case_when(
      data_type %in% pf_conf ~ "pf_conf",
      data_type %in% pv_conf ~ "pv_conf",
      data_type %in% positives ~ "positives",
      data_type %in% tested ~ "tested",
      data_type %in% clinical_malaria ~ "clinical",
      data_type %in% mixed_conf ~ "mixed_conf",
      data_type %in% po_conf ~ "po_conf",
      data_type %in% pm_conf ~ "pm_conf",
      data_type %in% actual_reports ~ "actual_reports",
      data_type %in% actual_on_time ~ "actual_reports_ontime",
      data_type %in% expected ~ "expected_reports",
      TRUE ~ data_type
    )
  )
```

### 2.2.5 Temporal standardization (Ethiopian to Gregorian calendar)

The period variable was parsed to extract Ethiopian calendar components and to derive corresponding Gregorian calendar dates. This was done to enable flexible time-series analysis while retaining the original Ethiopian calendar format for reference and reporting consistency.

Both calendar systems were retained in the final dataset to support different analytical and reporting needs.

```{r}
eth_months <- c("Meskerem","Tikemet","Hidar","Tahesas",
                "Tir","Yekatit","Megabit","Miazia",
                "Ginbot","Sene","Hamle","Nehase")

greg_months <- c(9,10,11,12,1,2,3,4,5,6,7,8)
month_map <- setNames(1:12, eth_months)

combined_hmis <- combined_hmis %>%
  mutate(
    eth_month = str_extract(period, "^[^ ]+"),
    eth_year  = as.integer(str_extract(period, "\\d+$")),
    eth_month_num = month_map[eth_month],
    greg_year = if_else(eth_month_num <= 4, eth_year + 7, eth_year + 8),
    greg_month = greg_months[eth_month_num],
    greg_date = make_date(greg_year, greg_month, 1)
  ) |>
  mutate(
    greg_month = month(greg_date, label = TRUE),
    greg_year = year(greg_date),
    month_year = paste(greg_month, "-", greg_year)
  ) |>
  select(
    region, zone, woreda,
    eth_month, eth_year,
    greg_date, greg_month, greg_year, month_year,
    facility_type, department, outcome, data_type, value
  )
```

The dataset was saved after general cleaning and harmonization for use in subsequent steps, including geographic cleaning and downstream analysis.

```{r}
saveRDS(
  combined_hmis,
  "data/processed/april 2026/apr_2026_hmis_general_cleaning_completed.rds"
)
```

## 2.3 Geographic name Cleaning

Following general data cleaning and harmonization, geographic variables were further standardized to ensure alignment between HMIS administrative unit names and the official Ethiopian administrative boundary reference.

Geographic cleaning was performed separately for each administrative level (region, zone, and woreda) using a standardized cleaning function developed by the PATH/MACEPA data analytics team. This function was used to harmonize geographic identifiers across datasets and align them with the reference administrative structure.

This step was implemented prior to indicator generation and visualization and was applied sequentially across all three administrative levels.

**Management of unmatched geographic names**

The geographic cleaning functions compare HMIS administrative unit names against the official administrative boundary reference dataset and the corresponding preprocessed correction files. In cases where unmatched names are identified, the unmatched records are exported for manual review.

These records are then assessed and mapped to the appropriate administrative unit names. The resulting corrections are added to the relevant preprocessing file (e.g., region, zone, or woreda correction file).

Once the correction file has been updated, the cleaning function can be rerun to apply the new mappings automatically. This process creates a growing repository of validated name corrections, reducing manual effort in future dashboard updates and improving consistency across reporting periods.

### 2.3.1 Region-level geographic cleaning

Region-level geographic cleaning was performed on the dataset resulting from general cleaning and harmonization. The objective was to standardize region names to match the official Ethiopian administrative reference.

This step ensured that all region identifiers in the HMIS dataset were consistent with the official administrative boundary definitions derived from the national shapefile reference.

*Inputs to the region cleaning process*

Three inputs were used in the region-level cleaning procedure:

- **HMIS dataset (input data):**\
  The output from general cleaning and harmonization\
  (`apr_2026_hmis_general_cleaning_completed.rds`)

- **Preprocessed region name corrections:**\
  A manually prepared mapping file containing inconsistent region names and their corresponding corrected values\
  (`region_names_corrected.csv`)

- **Reference region names:**\
  official administrative boundary reference dataset\
  (`eth-admin3-v1082-clean-id.rds`)

**Region cleaning process**

Region cleaning was performed using a standardized function developed by the PATH/MACEPA data analytics team. This function applies predefined matching logic, correction mappings, and reference validation to align region names with the official administrative structure.

```{r}
# libraries
library(tidyverse)
library(lubridate)
library(janitor)

# sourcing the geographic cleaning function
source("functions/geog_cleaning_accessories.R")

# paths for the files needed 
ref_path <- "data/eth-admin3-v1082-clean-id.rds"
dat_path <- "data/processed/april 2026/apr_2026_hmis_general_cleaning_completed.rds"
pre_path <- "data/preprocessed/region_names_corrected.csv"
out_path <- "data/processed/april 2026/apr_2026_hmis_region_cleaning_completed.rds"

# applying region cleaning function
rclean <- region_clean(
  dat_path = dat_path,
  pre_path = pre_path,
  ref_path = ref_path,
  out_path = out_path
)
```

- **Output of region cleaning**

The output of this step is a region-standardized HMIS dataset in which all region names are aligned with the official administrative reference.

This dataset serves as the input for subsequent zone-level geographic cleaning.

### 2.3.2 Zone-level geographic cleaning

Following region name standardization, zone names were cleaned and standardized to ensure alignment with the official administrative boundary reference dataset used in the dashboard.

This step was performed after region cleaning because zone matching relies on a consistent regional structure.

*Inputs to the zone cleaning process*

Three inputs were used in the zone-level cleaning procedure:

- **HMIS dataset (input data):**\
  Output from the region cleaning step\
  (`apr_2026_hmis_region_cleaning_completed.rds`)

- **Preprocessed zone name corrections:**\
  A manually prepared mapping file containing inconsistent zone names and their corresponding corrected values\
  (`zone_names_corrected.csv`)

- **Reference administrative boundary dataset:**\
  Official administrative names used as the reference for geographic standardization\
  (`eth-admin3-v1082-clean-id.rds`)

**Zone cleaning process**

Zone cleaning was performed using the standardized function developed by the PATH/MACEPA data analytics team. The function applies predefined correction mappings and reference matching procedures to align zone names with the official administrative hierarchy.

```{r}
# libraries
library(tidyverse)
library(lubridate)
library(janitor)

# sourcing the geographic cleaning function
source("functions/geog_cleaning_accessories.R")

# path for the reference administrative boundary dataset
ref_path <- "data/eth-admin3-v1082-clean-id.rds"

# path for the input dataset
dat_path <- "data/processed/april 2026/apr_2026_hmis_region_cleaning_completed.rds"

# path for preprocessed zone corrections
pre_path <- "data/preprocessed/zone_names_corrected.csv"

# path for the output dataset
out_path <- "data/processed/april 2026/apr_2026_hmis_zone_cleaning_completed.rds"

# clean zone names
zclean <- zone_clean(
  dat_path,
  pre_path,
  ref_path,
  out_path
)
```

### 2.3.3 Woreda-level geographic cleaning

Following zone-level standardization, woreda names were cleaned and standardized to ensure alignment with the official administrative boundary reference.

This step was performed after region and zone cleaning because woreda matching relies on a consistent administrative hierarchy. Standardizing woreda names ensures that HMIS records can be accurately linked to the corresponding administrative boundaries used for analysis and visualization.

*Inputs to the woreda cleaning process*

Three inputs were used in the woreda-level cleaning procedure:

- **HMIS dataset (input data):**\
  Output from the zone cleaning step\
  (`apr_2026_hmis_zone_cleaning_completed.rds`)

- **Preprocessed woreda name corrections:**\
  A cumulative correction file containing previously validated mappings between unmatched HMIS woreda names and the corresponding official administrative names\
  (`woreda_names_corrected.csv`)

- **Reference administrative boundary dataset:**\
  Official administrative names used as the reference for geographic standardization\
  (`eth-admin3-v1082-clean-id.rds`)

**Woreda cleaning process**

Woreda cleaning was performed using the same standardized function developed by the PATH/MACEPA team. The function applies predefined correction mappings and reference matching procedures to align woreda names with the official administrative hierarchy.

```{r}
# libraries 
library(tidyverse) 
library(lubridate) 
library(janitor)

# sourcing the geographic cleaning function
source("functions/geog_cleaning_accessories.R")

# path for the reference administrative boundary dataset
ref_path <- "data/eth-admin3-v1082-clean-id.rds"

# path for the input dataset
dat_path <- "data/processed/april 2026/apr_2026_hmis_zone_cleaning_completed.rds"

# path for preprocessed woreda corrections
pre_path <- "data/preprocessed/woreda_names_corrected.csv"

# path for the output dataset
out_path <- "data/processed/april 2026/apr_2026_hmis_woreda_cleaning_completed.rds"

# clean woreda names
wclean <- woreda_clean( dat_path, pre_path, ref_path, out_path )

```

**Integration of administrative identifiers**

After woreda names were standardized, the cleaned HMIS dataset was linked to the administrative boundary reference dataset to incorporate the unique woreda identifier (`id_1082`).

The `id_1082` variable is a standardized administrative code assigned to each woreda in the national administrative boundary dataset. Incorporating this identifier provides a stable geographic reference that is not affected by spelling variations or naming inconsistencies.

The identifier is used throughout the dashboard workflow for geographic aggregation, linkage with auxiliary datasets, and spatial visualization.

```{r}
# read cleaned HMIS data
clean_woredas <- readRDS(out_path) |>
  select(
    -region_old,
    -zone_old,
    -woreda_old,
    -non_zero_ever,
    -matched
  )

# read reference administrative boundary dataset
sf <- readRDS(ref_path) |>
  sf::st_drop_geometry() |>
  select(-region_old, -pop_2022)

# join administrative identifier
final_out <- clean_woredas |>
  left_join(
    sf,
    by = c("region", "zone", "woreda")
  )

# save final geographic-cleaned dataset
saveRDS(final_out, out_path)
```

**Output of woreda cleaning**

The output of this step is a dataset with standardized region, zone, and woreda names aligned with the official administrative boundary reference dataset.

This dataset serves as the geographic-cleaned HMIS dataset and provides the basis for subsequent aggregation, indicator calculation, visualization, and dashboard development.

# 3. dashboard-ready dataset preparation

## 3.1 Integration with the historical HMIS dataset

The dashboard is designed to display malaria trends over multiple years. Therefore, each newly processed monthly HMIS dataset must be integrated with the existing historical HMIS dataset maintained for the dashboard.

At this stage, the processed HMIS dataset for the current reporting period was appended to the historical HMIS dataset. This process creates an updated master dataset containing all available HMIS records required for dashboard analysis and visualization.

```{r}
# libraries
library(tidyverse)
library(lubridate)

# read historical HMIS dataset
hmis_large <- readRDS(
  "data/processed/jan 2020-mar 2026/clean_hmis_2020_2026.rds"
)

# read newly processed HMIS dataset
hmis_apr_2026 <- readRDS(
  "data/processed/april 2026/clean_apr_2026.rds"
)

# append new data to historical dataset
binded_hmis <- bind_rows(
  hmis_large,
  hmis_apr_2026
)

# save updated master dataset
saveRDS(
  binded_hmis,
  "data/processed/april 2026/upto_apr_2026_hmis_for_app.rds"
)
```

**Output of dataset integration**

The output of this step is an updated master HMIS dataset containing both the historical records and the newly processed reporting period.

This dataset serves as the source dataset for subsequent aggregation and indicator calculation steps used in the dashboard.

## 3.2 Creation of lookup tables and analytical dataset

Following geographic cleaning, a set of supporting lookup tables were prepared to facilitate dashboard functionality and indicator calculation.

The dashboard uses two types of lookup tables:

- **Static lookup tables**, which are created once and reused across routine dashboard updates. These include the population lookup and time lookup tables.

- **Dynamic lookup tables**, which are generated from the latest HMIS data and therefore require updating whenever a new HMIS extraction is processed. This includes the geographic lookup table.

These lookup tables are subsequently used to enrich the cleaned HMIS dataset and create the final analysis-ready dataset used throughout the dashboard.

**Geographic lookup table**

A geographic lookup table was generated from the geographically cleaned HMIS dataset. The lookup contains unique administrative identifiers and names for each region, zone, and woreda represented in the data.

Unlike the population and time lookup tables, the geographic lookup is derived directly from the HMIS dataset and should be regenerated whenever a new HMIS extraction is processed.

```{r}
#libraries
library(tidyverse)
library(lubridate)
library(sf)

# reading the cleaned hmis data
hmis <- readRDS("data/processed/april 2026/april_mar_2026_hmis_woreda_cleaning_completed.rds")

# read sf
sf <- readRDS("data/eth-admin3-v1082-clean-id.rds")|>
  sf::st_drop_geometry()


# prepare a geo look up
geo_lookup <- hmis %>%
  distinct(
    id_1082,
    region,
    zone,
    woreda
  ) %>%
  arrange(region, zone, woreda)

# save the geo look up
saveRDS(geo_lookup, "data/processed/april 2026/geo_lookup.rds")
```

**Population lookup table**

A population lookup table containing annual woreda-level population estimates was prepared for use in population-based indicator calculations.

This lookup is maintained separately from the HMIS data and can be reused across routine dashboard updates.

```{r}
# create a population lookup
pop_lookup <- read_csv("data/eth-admin3-v1082-clean-id_with_pop_2030.csv")|>
  filter(year %in% c(2020:2026))

# save population lookup (saved once, no need to repeat)
#saveRDS(pop_lookup, "data/processed/population_lookup.rds")
```

**Time lookup table**

A time lookup table was created to provide standardized Gregorian and Ethiopian Fiscal Year (EFY) date variables used throughout the dashboard.

This lookup is created independently of the HMIS data and can be reused across routine dashboard updates. It only requires modification if the dashboard period is extended beyond the range covered by the existing lookup table.

```{r}
# creating time lookup
time_lookup <- tibble(
  greg_date = seq(
    from = as.Date("2020-01-01"),
    to = as.Date("2030-12-01"),
    by = "month"
  )
)

# create gregorian variable
time_lookup <- time_lookup %>%
  mutate(
    greg_year = year(greg_date),
    
    greg_month = month(
      greg_date,
      label = TRUE,
      abbr = TRUE
    ),
    # character type gregorian year for dashboard UI
   greg_year_ui = factor(
        greg_year,
        levels = 2020:2026
      ),
    
    greg_month_num = month(greg_date),
    
    month_year = format(
      greg_date,
      "%b-%Y"
    )
  )

# create EFY variable
time_lookup <- time_lookup %>%
  mutate(
    
    efy = ifelse(
      greg_month_num >= 7,
      greg_year - 7,
      greg_year - 8
    ),
    
    efy_month_num = case_when(
      greg_month_num >= 7 ~ greg_month_num - 6,
      TRUE ~ greg_month_num + 6
    )
  )


# creating EFY display label
time_lookup <- time_lookup %>%
  mutate(
    efy_time = month_year
  )

# convert to factor
time_lookup <- time_lookup %>%
  mutate(
    
    greg_month = factor(
      greg_month,
      levels = c(
        "Jan","Feb","Mar","Apr",
        "May","Jun","Jul","Aug",
        "Sep","Oct","Nov","Dec"
      )
    ),
    
    month_year = factor(
      month_year,
      levels = format(
        greg_date,
        "%b-%Y"
      )
    )
  )

# save the time lookup (saved once, no need to repeat)
#saveRDS(time_lookup,"data/processed/time_lookup.rds")
```

## 3.3 Temporal enrichment of the HMIS dataset

After creating the time lookup table, it was joined to the geographically cleaned and integrated HMIS dataset using the Gregorian date variable (`greg_date`).

This step enriches the HMIS data with additional time-related variables required for dashboard filtering, aggregation, and visualization. These include Gregorian year and month variables, Ethiopian Fiscal Year (EFY) variables, and user-interface-friendly date fields.

By storing these variables in the HMIS dataset, time transformations do not need to be recalculated repeatedly during dashboard execution, improving efficiency and ensuring consistent date handling throughout the application.

```{r}
# join time lookup to the geography cleaned hmis
hmis <- hmis %>%
  left_join(
    time_lookup,
    by = "greg_date"
  )

# selecting the columns needed for subsequent processing
hmis <- hmis |>
  select(
    id_1082,
    region,
    zone,
    woreda,
    greg_date,
    greg_year = greg_year.y,
    greg_month = greg_month.y,
    greg_month_num,
    month_year = month_year.y,
    efy,
    greg_year_ui,
    efy_month_num,
    efy_time,
    facility_type,
    department,
    outcome,
    data_type,
    value
  )
```

**Output of lookup table creation and temporal enrichment**

The output of this step is a geographically standardized and temporally enriched HMIS dataset.

The dataset contains harmonized malaria indicators together with standardized administrative identifiers and Gregorian/Ethiopian Fiscal Year (EFY) time variables required for subsequent processing.

This dataset serves as the input for the aggregation step.

## 3.4 Aggregation

To improve dashboard performance and reduce computational load in the Shiny application, the integrated HMIS dataset was aggregated prior to deployment.

This step reduces data redundancy by summarizing records at the lowest analytical level required for visualization while preserving all key analytical dimensions.

```{r}
# libraries
library(tidyverse)
library(lubridate)
library(sf)

# read integrated HMIS dataset
hmis <- readRDS(
  "data/processed/april 2026/upto_apr_2026_hmis_for_app.rds"
)

# aggregate HMIS data for dashboard efficiency
hmis_dashboard <- hmis |>
  group_by(
    id_1082,
    region,
    zone,
    woreda,
    facility_type,
    greg_date,
    department,
    outcome,
    data_type
  ) |>
  summarise(
    value = sum(value, na.rm = TRUE),
    .groups = "drop"
  )
```

**Output of aggregation**

The output of this step is a reduced and performance-optimized dataset (hmis_dashboard) in which duplicate records have been aggregated while preserving all analytical dimensions required for dashboard visualizations.

This improves dashboard responsiveness by reducing memory usage and computational overhead during filtering and rendering.

# 4. Visualization and Dashboard Development

## 4.1 Application initialization and data loading

Before launching the user interface and server components, the application loads all required datasets, lookup tables, geographic boundary files, and supporting objects.

This initialization step ensures that all data required for filtering, indicator calculation, and visualization are available when the dashboard starts.

The following resources are loaded:

- Dashboard-ready HMIS dataset
- Population lookup table
- Geographic lookup table
- Administrative boundary shapefile

In addition, geographic layers are prepared for woreda-, zone-, and region-level mapping, and application-wide settings such as date ordering and visual themes are initialized.

## 4.2 Dashboard structure and navigation

The dashboard was developed as a web-based interactive application using R Shiny. The application is organized into two main pages designed to support both user orientation and data exploration.

**Overview page**

The Overview page serves as the entry point to the dashboard and provides background information to help users understand the purpose, scope, and interpretation of the dashboard outputs.

The page includes the following sections:

- About the dashboard
- Data sources
- Key indicators
- Methodological notes and
- Dashboard user guide

This page is intended to provide contextual information and support consistent interpretation of dashboard outputs.

**Dashboard page**

The Dashboard page contains all interactive visualizations and analytical outputs. It is organized into a sidebar panel containing user-defined filters and a main panel displaying a map, key performance indicators (KPIs), and charts.

**Sidebar filters**

The sidebar contains three groups of filters:

*Geographic filters*

- Region
- Zone
- Woreda

These filters allow users to view data at different administrative levels and geographic locations.

*Time filters*

- Gregorian year
- Ethiopian Fiscal Year (EFY)
- Month

These filters enable temporal analysis of malaria indicators.

*Reporting filters*

- Facility type

The facility type filter is used only for reporting completeness and timeliness indicators. All other filters apply across all dashboard components.

**Dashboard outputs**

The main dashboard panel is organized into four rows of analytical outputs.

*Row 1*

- Geographic map
- Key Performance Indicators (KPIs)

*Row 2*

- Malaria testing and Test Positivity Rate (TPR)
- Species proportion

The testing and TPR visualization combines malaria testing volume as a bar chart with TPR displayed as a line on a secondary axis.

*Row 3*

- Parasite incidence per 1,000 population
- Malaria mortality rate per 100,000 population at risk

*Row 4*

- Reporting completeness
- Reporting timeliness

**Visualization modes**

All dashboard charts include a display toggle that allows users to switch between:

- Aggregate view and
- Monthly trend view

The aggregate view summarizes indicator values over the selected period and geography, while the monthly view displays temporal trends across individual months.

# 5. Indicator Definitions and Calculations

The dashboard presents key malaria surveillance, testing, mortality, species distribution, and reporting performance indicators derived from HMIS data. Indicator definitions and calculation methods are described below.

## 5.1 Malaria Cases

**Confirmed Malaria Cases**

Confirmed malaria cases include all laboratory-confirmed malaria infections reported through HMIS, regardless of species.

**Clinical Malaria Cases**

Clinical malaria cases refer to malaria diagnoses made without parasitological confirmation.

**All Malaria Cases**

All malaria cases represent the total malaria burden reported through HMIS and are calculated as:

*All Cases = Confirmed Cases + Clinical Cases*

## 5.2 Malaria Deaths

Malaria deaths are defined as inpatient malaria-related deaths reported through the HMIS disease surveillance system.

## 5.3 Test Positivity Rate (TPR)

The Test Positivity Rate (TPR) measures the proportion of malaria tests that yield a positive result.

*TPR = Positive Malaria Tests/Total malaria test performed x 100*

Where:\
- Positive malaria tests = laboratory-confirmed positive malaria tests\
- Total malaria tests performed = all malaria diagnostic tests (microscopy and/or RDT)

## 5.4 Species Distribution

Species proportions describe the relative contribution of each malaria species among confirmed malaria cases.

**Plasmodium falciparum Proportion** = Pf cases/Total confirmed cases\*100

**Plasmodium vivax Proportion** = Pv cases/Total confirmed cases\*100

**Mixed malaria (pf/pv) Proportion** = Mixed cases/Total confirmed cases\*100

*Total Confirmed Cases = Pf + Pv + Mixed + P. malariae + P. ovale*

## 5.5 Annual Parasite Incidence (API)

The dashboard uses adjusted malaria cases to account for clinically diagnosed malaria cases.

Adjusted malaria cases are calculated as:

*Adjusted Cases = Confirmed Cases + (TPR × Clinical Cases)*

Annual Parasite Incidence (API) is then calculated as:

*API = Adjusted Cases/ Total Population × 1000* and expressed as cases per 1,000 population.

## 5.6 Annualized Monthly Parasite Incidence

For monthly dashboard displays, incidence is annualized to facilitate comparison with annual incidence targets.

*Annualized monthly incidence = (Adjusted cases X n/12)/Total population X 1000*

Where:\
*n = annualization factor used to project the observed reporting period to a 12-month equivalent*

## 5.7 Malaria Mortality Rate

The malaria mortality rate is calculated using the population at risk as the denominator.

In accordance with the National Malaria Strategic Plan,the population at risk is defined as:  *Population at Risk = Total Population × 69%*

The mortality rate is calculated as:

*Motality_rate = In-patient malaria deaths/population at risk X 100,000* and expressed as deaths per 100,000 population at risk.

## 5.8 Annualized Monthly Mortality Rate

For monthly dashboard displays, mortality is annualized using the same approach applied to incidence.

*Annualized monthly mortality = (inpatient malaria deaths X n/12)/population at risk X 100,000*

Where:\
*n = annualization factor used to project the observed reporting period to a 12-month equivalent*

## 5.9 Reporting Completeness

Reporting completeness measures the proportion of expected reports that were submitted.

*Reporting completeness = Actual reports/Expected reports X 100*

## 5.10 Reporting Timeliness

Reporting timeliness measures the proportion of expected reports that were submitted on time.

*Reporting timeliness = Reports received on time/Expected reports X 100*

**Notes on Reporting Indicators**

Reporting completeness and reporting timeliness are assessed using only the **HMIS service reporting dataset**. Both the numerator and denominator values are system-generated by DHIS2 and are not calculated manually within the dashboard.

These indicators assess reporting performance rather than malaria burden and are displayed separately from surveillance and service delivery indicators.

# 6. Monthly Dashboard Update Procedure

The dashboard is designed to be updated monthly as new HMIS data become available. The update process involves downloading the latest HMIS data, running the existing data processing scripts, replacing the dashboard dataset, and republishing the application.

Near-complete HMIS data are typically available in DHIS2 from approximately the 22nd day of the following month. Dashboard updates should therefore be performed after this date to ensure adequate reporting completeness.

**The monthly update workflow consists of the following steps:**

1.  Download the latest HMIS datasets from DHIS2:
    - Disease surveillance dataset

    - Service delivery dataset

    - Reporting completeness and timeliness dataset

      Save the downloaded files in the designated raw data folder.
2.  Run the data processing scripts sequentially:
    - Data preparation script

    - General data cleaning and harmonization script

    - Region-level geographic cleaning script

    - Zone-level geographic cleaning script

    - Woreda-level geographic cleaning script

    - Dashboard dataset preparation script

    - Historical data integration script

    - Dashboard aggregation script
3.  Review any unmatched geographic names generated during the cleaning process.
    - Update the corresponding correction files if new unmatched names are identified:

      - region_names_corrected.csv

      - zone_names_corrected.csv

      - woreda_names_corrected.csv
4.  Rerun the relevant geographic cleaning script(s) if correction files are updated.
5.  Generate the updated dashboard-ready dataset.
6.  Replace the existing dashboard dataset with the newly generated dataset.
7.  Run the Shiny application and verify that:
    - The latest reporting month is available.

    - Dashboard visualizations render correctly.

    - Key indicators and maps display expected values.

    - Filters function as intended.
8.  Republish the dashboard to ShinyApps.io. Perform a final review of the published dashboard to confirm that the updated data are visible and functioning correctly.

This workflow should be repeated each month to maintain an up-to-date dashboard for routine malaria surveillance and program monitoring.
